Nonlinear methods to quantify Movement Variability in Human-Humanoid Interaction Activities

### 2021-03-18 docs/manuscript/main.tex
ammends affiliations and correspondance email
figure01 harmonised and updated

\title{Nonlinear methods to quantify Movement Variability in Human-Humanoid Interaction Activities}
\author[1,*]{Miguel Xochicale}
\author[2]{Chris Baber}
\affil[1]{King's College London, 
	School of Biomedical Engineering amd Imaging Sciences,
	London, SE1 7EU, UK
	} 
\affil[2]{University of Birmingham,
	School of Computer Science,
	Birmingham, 
	B15 2TT, 
	UK}
\affil[*]{miguel.xochicale@kcl.ac.uk}

@@

Figures~\ref{fig01}(A.1 to D.1) illustrates the state space reconstruction.

@@

\includegraphics[width=1.0\textwidth]{figures/fig01/versions/drawing-v00}

@@

	{\bf State space reconstruction methodology.}

@@

    	(A.1) $M-$dimensional manifold representing the state space (e.g. Lorenz system);
    	(B.1) Delayed copies of $1-$dimensional $x(t)$ from the Lorenz system;
		(C.1) $m-$dimensional reconstructed state space with \texorpdfstring{$m$}{m} and    \texorpdfstring{$\tau$}{T}, and
    	(D.1) $y(t)$ is the $n-$dimensional reconstructed state space.

@@

	A.1 to D.1 figures are adapted from \cite{Quintana-Duque2012, casdagli1991, uzal2011}.
	{\bf Minimum delay embedding values with AMI's method and minimum dimension embedding values with Cao's method.} 
	(A.2) chaotic and (B.2) random time series.
	(C.2, D.2) $E_1 (m)$ values and (E.2, F.2) $E_2(m)$ values with variations of $\tau$ values from one to twenty for chaotic and random time series.
	(G.2, H.2) AMI values where its first minimum value in the curve is the minimum time delay embedding ($\tau_0$).
	{\bf Recurrence Plots and its patterns.} 
	(A.3) State space of the Lorenz system with controlling parameters ($\rho=28, \sigma=10, \beta=8/3$) where the point, $j$, in trajectory $X()$ which falls into the neighborhood (black circle) of a given point at $i$ is a recurrent point and is represented as a black dot in the recurrence plot at location $(i, j)$ or white otherwise.
	(B.3) Recurrence plot using the three components of the Lorenz system and the RP with no embeddings and threshold $\epsilon=5$.

@@

	(C.3) uniformly distributed noise,
	(D.3) super-positionet harmonic oscillation ($sin( \frac{1}{5}*t) * sin( \frac{5}{100}*t) $),
	(E.3) drift logistic map ($x_{i+1} = 4 x_i (1- x_i) $) corrupted with a linearly increase term ($0.01*i$), and
	(F.3) disrupted brownian motion  ($x_{i+1} = x_i + 2*rnorm(1) $).
	Figures A.3 to F.3 are adapted from \cite{marwan2015}.

@@

As an example of the use of $E_1(m)$ and $E_2(m)$ values, we consider two time 
series: the solution for the $x$ variable of the Lorenz system 
(Fig~\ref{fig01}A.2), and a Gaussian noise time series with zero mean
and a variance of one (Fig~\ref{fig01}B.2).
We then compute $E_1(m)$ and $E_2(m)$ values for each time series.
The $E_1(m)$ values for the chaotic time series appear to be constant
after the dimension is equal to six.
The determination of six is given that any value of $m$ can be used as they 
are within the threshold of $1\pm0.05$ (Fig~\ref{fig01}C.2).
$E_2(m)$ values, for chaotic time series, are different to one 
(Fig~\ref{fig01}E.2), for which, it can be concluded that for the
chaotic time series the minimum embedding dimension the time series 
comes from a deterministic signal. With regard to the noise time series,  
$E_1(m)$ values appeared to be constant when $m$ is close to thirteen, 
which is defined by the threshold of $1\pm0.05$ (Fig~\ref{fig01}D.2).
$E_1(m)$ values then indicate the minimum embedding dimension of the 
noisy time series is thirteen, however all of the $E_2(m)$ values are 
approximately equal to one (Figure~\ref{fig01}F.2) for which it can be
concluded that noise time series is a stochastic signal.
It is important to note that for this work not only $E_1(m)$ and $E_2(m)$ are 
computed but also a variation of $\tau$ from 1 to 20 is presented. 
The purpose of such variation for $\tau$ is to show its independence with
regard to $E_1(m)$ and $E_2(m)$ values as $\tau$ is increasing 
(Fig~\ref{fig01}C.2,D.2,E.2 and F.2).
However, one negative of the Cao's algorithm \cite{Cao1997} is the definition of 
a new threshold where $m$ values appear to be constant in $E_1 (m)$.
In the case of the given examples and reported results, we defined such 
threshold as 0.05. Further investigation is required for the selection of the 
threshold in the $E_1(m)$, as the selection of the threshold in this work is 
base on no particular method but visual inspection.

@@

For example, we compute the AMI for two time series:
A) the $x$ solution of the Lorenz system, and
B) a noise time series using a normal distribution with mean zero and 
standard deviation equal to one. From Fig~\ref{fig01}(G.2, H.2), it can then be
concluded that the amount of knowledge for any noise time series is zero 
for which the first minimum embedding parameter is $\tau_0=1$. 
On the contrary, the first minimum of the AMI for the chaotic time series 
is $\tau_0=17$ which is the value that maximize the independence 
between $x(n)$ and $x(n+\tau)$ in the reconstructed state space 
\cite{bradley2015}.
%%---------------------------------(FIGURE)-------------------------------------
Similarly as Cao's algorithm negatives, AMI's algorithm is not an
exception for negatives, which are worthwhile to mention for further 
investigations.
For instance, (i) is not clear why the choose of the first minimum of the AMI 
is the minimum delay embedding parameter \cite{kantz2003} and 
(ii) the probability distribution of the AMI function is computed
with the use of histograms which depends on a heuristic choice of number of bins
for which AMI depends on partitioning \cite{garcia2005e71}.

@@

For the creation of a recurrence plot based on time series 
$\{ \boldsymbol{x}_n \}$, it is first computed the state space 
reconstruction with uniform time-delay embedding 
$X(i)=\{ \boldsymbol{ \tilde{x} }_n, \dots,  
\boldsymbol{ \tilde{x} }_{n -(m-1)\tau} \}$
where $i=1,\dots,N$, $N$ is the number of considered states of $X(i)$ 
and $X(i) \in \mathbb{R}^m$ \cite{eckmann1987}.
The recurrence plot is therefore a two-dimensional $N \times N$ square 
matrix, $\mathbf{R}$, where a black dot is placed at $(i,j)$ 
whenever $X(i)$ is sufficiently close to $X(j)$: 
%%********************************[EQUATION]************************************
\begin{equation}
\mathbf{R}^{m}_{i,j} (\epsilon) = \Theta ( \epsilon_i - || X(i) - X(j) ||
\end{equation}
%%********************************[EQUATION]************************************
where $\quad i,j=1,\dots,N$, $\epsilon$ is a threshold distance, 
$|| \cdotp ||$ a norm, and $\Theta(\cdotp)$ is the Heaviside 
function (i.e. $\Theta(x)=0$, if $x<0$, and $\Theta(x)=1$ otherwise) 
(Fig~\ref{fig01}(A.3,B.3)) \cite{eckmann1987, marwan2007,marwan2015}.
RP is also characterised with a line of identity (LOI) which is a  
diagonal line due to $ R_{i,j}=1 (i,j=1,\dots,N)$.

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\subsection*{Structures of Recurrence Plots}
Pattern formations in the RPs can be designated either 
as topology for large-scale patterns or texture for small-scale patterns.
In the case of topology, the following pattern formations are commonly presented:
(i) homogeneous where uniform recurrence points are spread in the RP e.g., 
uniformly distributed noise (Fig~\ref{fig01}C.3),
(ii) periodic and quasi-periodic systems where diagonal lines and 
checkerboard structures represent oscillating systems, e.g., sinusoidal 
signals (Fig~\ref{fig01}D.3),
(iii) drift where paling or darkening recurrence points away from 
the LOI is caused by drifting systems, 
e.g., logistic map (Fig~\ref{fig01}E.3), and
(iv) disrupted where recurrence points are presented white areas or 
bands that indicate abrupt changes in the dynamics, e.g. Brownian motion 
(Fig~\ref{fig01}F.3) \cite{eckmann1987, marwan2015}.
Texture patterns in RPs can be categorised as:
(i) single or isolated recurrence points that represent rare occurring states, 
and do not persist for any time or fluctuate heavily,
(ii) dots forming diagonal lines where the length of the small-scale parallel 
lines in the diagonal are related to the ratio of determinism or predictability 
in the dynamics of the system, and
(iii) dots forming vertical and horizontal lines where the length of the 
lines represent a time length where a state does not change or change very 
slowly and these patterns formation represent discontinuities in the signal, 
and (iv) dots clustering to inscribe rectangular regions which are related 
to laminar states or singularities \cite{marwan2015}.

>>>>

\title{Nonlinear methods to quantify Movement Variability in Human-Humanoid Interaction Activities}
\author[1,*]{Miguel Xochicale}
\author{Chris Baber}
\affil[1]{University of Birmingham,
	School of Computer Science,
	Birmingham, 
	B15 2TT, 
	UK}
\affil[*]{perez.xochicale@gmail.com}

@@

Figures~\ref{fig01}A illustrates the state space reconstruction.

@@

\includegraphics[width=1.0\textwidth]{figures/fig01/versions/drawing-v01}

@@

	{\bf State space reconstruction methodology (A),
		Minimum delay embedding values with AMI's method and minimum dimension embedding values with Cao's method (B), and
		Recurrence Plots and its patterns (C).

@@

    	(a1) $M-$dimensional manifold representing the state space (e.g. Lorenz system);
    	(a2) Delayed copies of $1-$dimensional $x(t)$ from the Lorenz system;
		(a3) $m-$dimensional reconstructed state space with \texorpdfstring{$m$}{m} and    \texorpdfstring{$\tau$}{T}, and
    	(a4) $y(t)$ is the $n-$dimensional reconstructed state space.

@@

	a1 to a4 figures are adapted from \cite{Quintana-Duque2012, casdagli1991, uzal2011}.
	(b1) chaotic and (b5) random time series.
	(b2, b6) $E_1 (m)$ values and (b3, b7) $E_2(m)$ values with variations of $\tau$ values from one to twenty for chaotic and random time series.
	(b4, b8) AMI values where its first minimum value in the curve is the minimum time delay embedding ($\tau_0$).
	(c1) State space of the Lorenz system with controlling parameters ($\rho=28, \sigma=10, \beta=8/3$) where the point, $j$, in trajectory $X()$ which falls into the neighborhood (black circle) of a given point at $i$ is a recurrent point and is represented as a black dot in the recurrence plot at location $(i, j)$ or white otherwise.
	(c2) Recurrence plot using the three components of the Lorenz system and the RP with no embeddings and threshold $\epsilon=5$.

@@

	(c3) uniformly distributed noise,
	(c4) super-positionet harmonic oscillation ($sin( \frac{1}{5}*t) * sin( \frac{5}{100}*t) $),
	(c5) drift logistic map ($x_{i+1} = 4 x_i (1- x_i) $) corrupted with a linearly increase term ($0.01*i$), and
	(c6) disrupted brownian motion  ($x_{i+1} = x_i + 2*rnorm(1) $).
	Figures in C are adapted from \cite{marwan2015}.

@@

As an example of the use of $E_1(m)$ and $E_2(m)$ values, we consider two time 
series: the solution for the $x$ variable of the Lorenz system 
(Fig~\ref{fig01}b1), and a Gaussian noise time series with zero mean
and a variance of one (Fig~\ref{fig01}b5).
We then compute $E_1(m)$ and $E_2(m)$ values for each time series.
The $E_1(m)$ values for the chaotic time series appear to be constant
after the dimension is equal to six.
The determination of six is given that any value of $m$ can be used as they 
are within the threshold of $1\pm0.05$ (Fig~\ref{fig01}b2).
$E_2(m)$ values, for chaotic time series, are different to one 
(Fig~\ref{fig01}b3), for which, it can be concluded that for the
chaotic time series the minimum embedding dimension the time series 
comes from a deterministic signal. With regard to the noise time series,  
$E_1(m)$ values appeared to be constant when $m$ is close to thirteen, 
which is defined by the threshold of $1\pm0.05$ (Fig~\ref{fig01}b6).
$E_1(m)$ values then indicate the minimum embedding dimension of the 
noisy time series is thirteen, however all of the $E_2(m)$ values are 
approximately equal to one (Figure~\ref{fig01}b7) for which it can be
concluded that noise time series is a stochastic signal.
It is important to note that for this work not only $E_1(m)$ and $E_2(m)$ are 
computed but also a variation of $\tau$ from 1 to 20 is presented. 
The purpose of such variation for $\tau$ is to show its independence with
regard to $E_1(m)$ and $E_2(m)$ values as $\tau$ is increasing 
(Fig~\ref{fig01}b2,b3,b6, and b7).
However, one negative of the Cao's algorithm \cite{Cao1997} is the definition of 
a new threshold where $m$ values appear to be constant in $E_1 (m)$.
In the case of the given examples and reported results, we defined such 
threshold as 0.05. Further investigation is required for the selection of the 
threshold in the $E_1(m)$, as the selection of the threshold in this work is 
base on no particular method but visual inspection.

@@

For example, we compute the AMI for two time series:
A) the $x$ solution of the Lorenz system, and
B) a noise time series using a normal distribution with mean zero and 
standard deviation equal to one. From Fig~\ref{fig01}(b4, b8), it can then be
concluded that the amount of knowledge for any noise time series is zero 
for which the first minimum embedding parameter is $\tau_0=1$. 
On the contrary, the first minimum of the AMI for the chaotic time series 
is $\tau_0=17$ which is the value that maximize the independence 
between $x(n)$ and $x(n+\tau)$ in the reconstructed state space 
\cite{bradley2015}.
%%---------------------------------(FIGURE)-------------------------------------
Similarly as Cao's algorithm negatives, AMI's algorithm is not an
exception for negatives, which are worthwhile to mention for further 
investigations.
For instance, (i) is not clear why the choose of the first minimum of the AMI 
is the minimum delay embedding parameter \cite{kantz2003} and 
(ii) the probability distribution of the AMI function is computed
with the use of histograms which depends on a heuristic choice of number of bins
for which AMI depends on partitioning \cite{garcia2005e71}.

@@

For the creation of a recurrence plot based on time series 
$\{ \boldsymbol{x}_n \}$, it is first computed the state space 
reconstruction with uniform time-delay embedding 
$X(i)=\{ \boldsymbol{ \tilde{x} }_n, \dots,  
\boldsymbol{ \tilde{x} }_{n -(m-1)\tau} \}$
where $i=1,\dots,N$, $N$ is the number of considered states of $X(i)$ 
and $X(i) \in \mathbb{R}^m$ \cite{eckmann1987}.
The recurrence plot is therefore a two-dimensional $N \times N$ square 
matrix, $\mathbf{R}$, where a black dot is placed at $(i,j)$ 
whenever $X(i)$ is sufficiently close to $X(j)$: 
%%********************************[EQUATION]************************************
\begin{equation}
\mathbf{R}^{m}_{i,j} (\epsilon) = \Theta ( \epsilon_i - || X(i) - X(j) ||
\end{equation}
%%********************************[EQUATION]************************************
where $\quad i,j=1,\dots,N$, $\epsilon$ is a threshold distance, 
$|| \cdotp ||$ a norm, and $\Theta(\cdotp)$ is the Heaviside 
function (i.e. $\Theta(x)=0$, if $x<0$, and $\Theta(x)=1$ otherwise) 
(Fig~\ref{fig01}(c1,c2)) \cite{eckmann1987, marwan2007,marwan2015}.
RP is also characterised with a line of identity (LOI) which is a  
diagonal line due to $ R_{i,j}=1 (i,j=1,\dots,N)$.

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\subsection*{Structures of Recurrence Plots}
Pattern formations in the RPs can be designated either 
as topology for large-scale patterns or texture for small-scale patterns.
In the case of topology, the following pattern formations are commonly presented:
(i) homogeneous where uniform recurrence points are spread in the RP e.g., 
uniformly distributed noise (Fig~\ref{fig01}c3),
(ii) periodic and quasi-periodic systems where diagonal lines and 
checkerboard structures represent oscillating systems, e.g., sinusoidal 
signals (Fig~\ref{fig01}c4),
(iii) drift where paling or darkening recurrence points away from 
the LOI is caused by drifting systems, 
e.g., logistic map (Fig~\ref{fig01}c5), and
(iv) disrupted where recurrence points are presented white areas or 
bands that indicate abrupt changes in the dynamics, e.g. Brownian motion 
(Fig~\ref{fig01}c7) \cite{eckmann1987, marwan2015}.
Texture patterns in RPs can be categorised as:
(i) single or isolated recurrence points that represent rare occurring states, 
and do not persist for any time or fluctuate heavily,
(ii) dots forming diagonal lines where the length of the small-scale parallel 
lines in the diagonal are related to the ratio of determinism or predictability 
in the dynamics of the system, and
(iii) dots forming vertical and horizontal lines where the length of the 
lines represent a time length where a state does not change or change very 
slowly and these patterns formation represent discontinuities in the signal, 
and (iv) dots clustering to inscribe rectangular regions which are related 
to laminar states or singularities \cite{marwan2015}.
